``Interestingness'' in Mathematical Theory Exploration

### 2018-05-17 PhDThesis/background_theoryexploration.tex
More imports

>>>>

%%%%%

\begin{description}
\item{Interestingness} Various alternative interestingness criteria have been
  proposed, which we survey in \S \ref{sec:relatedwork}. Augmenting or replacing
  the criteria may be useful, for example to distinguish useful relationships
  from incidental coincidences; or to prevent surprising, insightful equations
  from being discarded because they can be simplified.
\end{description}

% Theory exploration is similar to \emph{experimental mathematics}

% - Relation to Science
%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
% - Relation to AI tasks: exploring surroundings, etc.

% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.

% Theory formation: Alison? Others.
% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
% Communication: Latex, Wikis, APIs, communicating with aliens

\section{Exploration in Theorem Proving}
\label{sec:examples}

Before exploring abstract definitions of interestingness, we can first consider
some scenarios which arise during formal proof where we are forced to generate
conjectures. An analysis of these situations, and the subsequent theorems they
produce, will contribute towards an empirical justification for what is
interesting (at least from a utilitarian point of view) and inform our later
exploration of the literature.

\subsection{Generalisation}

\providecommand{\coq}[1]{\lstinline[language=ML]|#1|}

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
which $S$ is a special case. Although it seems counterintuitive, a generalised
statement can sometimes be \emph{easier} to prove than the original. This arises
often in inductive proofs, since the specific obligations which arise in the
proof may be incompatible with the available inductive hypotheses.

However, we cannot blindly generalise \emph{all} obligations we encounter, since
\emph{over-generalising} results in obligations which are so strong that they
are unprovable, or even false. We must therefore rely on heuristics to guide the
generation of generalised conjectures, and hence perform a kind of exploration.

An informative example is given by Boyer and Moore of the associativity of
multiplication in ACL2 \cite{boyer1983proof}:

$$(x * y) * z = x * (y * z)$$

During the course of the proof, the following obligation arises:

\begin{equation}
  \tag{conc3}
  (y + (x * y)) * z = (y * z) + ((x * y) * z)
  \label{eq:conc3}
\end{equation}

ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated
sub-term $x * y$ with a fresh variable $w$:

\begin{equation}
  \tag{conc4}
  (y + w) * z = (y * z) + (w * z)
  \label{eq:conc4}
\end{equation}

This generalised form is clearly the distributivity law for multiplication and
addition, which can be proved separately to the original goal of
associativity. It would not be controversial to claim that this distributivity
law is interesting in its own right (relative to associativity, at least), in
addition to its usefulness in making this proof go through.

% TODO: Describe the ACL2 heuristics

Generalisation also occurs frequently when reasoning about \emph{tail-recursive}
definitions \cite{kapur2003automatic}. \footnote{A tail-recursive function can
  be executed in constant space using a loop, whereas recursion in non-tail
  positions may require a growing number of stack frames or nested closures. See
  \S \ref{sec:auxiliarylemmas} for example definitions of each type.}

\subsection{Analogy}

One way to characterise the interestingness of a statement is by \emph{analogy}
to existing interesting statements. By finding lemmas analogous to those of a
different theory, we may be able to re-use tactics and other forms of
meta-programming across both.

Existing theory exploration systems have been successfully applied to this
problem, however the use of pure exploration misses opportunities to
\emph{focus} the search, since we know which lemmas are used in those theories
where a technique succeeded. If we can find an analogy to map from such solved
problems to our unsolved goal, we can infer the approximate form of the lemmas
we require, and target these specifically.

The approach taken by \textsc{ACL2(ml)} is to find lemmas which may be relevant
to solving a goal $G$ by making analogies via unsupervised clustering
\cite{Heras.Komendantskaya.Johansson.ea:2013}. These clusters are used in two
ways:

\begin{itemize}
\item First, we use the cluster $C_G$ containing $G$ to identify analogous
  theorems.

\item For each theorem $T \in C_G \setminus \{G\}$, we consider those symbols
  $S_T$ which occur in $T$ but not in $G$. Our analogous lemmas are those used
  to prove $T$, mutated such that symbols $s \in S_T$ are replaced by members of
  the cluster $C_s$ containing $s$.
\end{itemize}

The running examples for demonstrating \textsc{ACL2(ml)} are equivalence
theorems for tail-recursive and non-tail-recursive calculations, as well as the
effect of repeating certain list operations:

\begin{itemize}

  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fact-tail}(n) = \texttt{fact}(n)$ where \texttt{natp} is the predicate that $n$ is a natural number, whilst \texttt{fact-tail} and \texttt{fact} are tail-recursive and non-tail-recursive implementations of factorial, respectively.

  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{power-tail}(n) = \texttt{power}(n)$,  where \texttt{power-tail} and \texttt{power} calculate powers of 2.

  \item $\forall n, \texttt{natp}(n) \rightarrow \texttt{fib-tail}(n) = \texttt{fib}(n)$,  where \texttt{fib-tail} and \texttt{fib} calculate fibonacci numbers.

  \item $\forall x, \texttt{nat-listp}(x) \rightarrow \texttt{sort}(\texttt{sort}(x)) = \texttt{sort}(x)$, for list-of-natural-numbers predicate \texttt{nat-listp} and list-sorting function \texttt{sort}.

  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{rev}(\texttt{rev}(x)) = x$, where \texttt{true-listp} ensures that $x$ is a valid singly-linked list structure and \texttt{rev} is list reversal.

  \item $\forall x, \texttt{true-listp}(x) \rightarrow \texttt{int}(x, x) = x$, where \texttt{int} is the intersection of lists (i.e. a list of elements common to each).

\end{itemize}

\subsection{Auxiliary Lemmas} \label{sec:auxiliarylemmas}

One consideration when generating conjectures is the difference between
theorems, lemmas, corollaries, etc. From a logical point of view, these are all
equivalent, and hence most proof assistants do not distinguish between
them. However, their \emph{intention} may be different: in a sense, theorems are
the interesting results; whilst lemmas are useful results, required for proving
the theorems.

Some systems, like Coq, allow users to \emph{label} each statement as being a
\coq{Theorem}, a \coq{Lemma}, etc. despite their internal representations being
the same. This shows us immediately that lemmas outnumber theorems; in the Coq
standard library there are over five times as many lemmas as theorems
\footnote{The latest version as of writing is \texttt{coq-8.4pl6} which, when
  excluding comments, includes 1492 occurences of \coq{Theorem} and 7594 of
  \coq{Lemma} in its \texttt{theories/} directory.}.

% TODO: Analyse them

% TODO: Theory exploration as lemma generation; give example from a HipSpec paper

We can find a need for auxiliary lemmas, once again, in the context of
tail-recursive functions. Consider proving the (pointwise) equality of the
following Coq functions, defined for the Peano naturals \coq{Z} and \coq{S}:

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
Inductive Nat : Set := Z : Nat
                     | S : Nat -> Nat.

Fixpoint plus      (n m : Nat) := match n with
                                      | Z    => m
                                      | S n' => S (plus n' m)
                                  end.

Fixpoint plus_tail (n m : Nat) := match n with
                                      | Z    => m
                                      | S n' => plus_tail n' (S m)
                                  end.
\end{lstlisting}

\begin{haskell}
Haskell equivalent:

plus :: Nat -> Nat -> Nat
plus      n  Z    = n
plus      n (S m) = S (plus n m)

plus_tail :: Nat -> Nat -> Nat
plus_tail n  Z    = n
plus_tail n (S m) = plus_tail (S n) m
\end{haskell}

Both of these functions implement addition, but the \coq{plus_tail} variant is
tail-recursive. However, if we want to \emph{prove} that the definitions are
(pointwise) equal, we run into difficulties. In particular, when the inductive
step requires us to prove \coq{plus (S n) m = plus n (S m)} (which seems
reasonable), we cannot make this go through using another inductive argument.

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
(* Solve equalities by beta-normalising both sides *)
Ltac triv := try (simpl; reflexivity).

(* Prove equivalence of plus and plus_tail *)
Theorem equiv : forall n m, plus n m = plus_tail n m.
  induction n; triv. (* Base case is trivial *)

  (* Inductive case: plus (S n) m = plus_tail (S n) m *)
  intro m.

  (* Beta-reduce the right-hand-side (justification is trivial) *)
  replace (plus_tail (S n) m) with (plus_tail n (S m)); triv.

  (* Use induction hypothesis to replace plus_tail with plus *)
  rewrite <- (IHn (S m)).
\end{lstlisting}

Specifically, the \emph{conclusion} of a second inductive hypothesis is exactly
the equation we need:

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
IHn' : (forall x, plus n' x = plus_tail n' x) -> plus (S n') m = plus n' (S m)
\end{lstlisting}

Yet we cannot provide it with the argument it needs, as our original induction
hypothesis is \emph{too specific} (i.e. it has too many \coq{S} constructors):

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
IHn : forall x, plus (S n') x = plus_tail (S n') x
\end{lstlisting}

We are forced to abandon the proof, despite such a reasonable-looking
intermediate goal.

In fact, if we attempt to prove that goal \emph{separately}, we can use a
straightforward argument by induction; even though it is actually
\emph{stronger} due to the absence of the \coq{IHn} assumption. Using this
separate result as a lemma, the pointwise equality is proven easily.

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
Lemma gen n m : plus (S n) m = plus n (S m).
  induction n; triv. (* Base case is trivial *)

  (* Move all S constructors outside *)
  simpl. rewrite <- IHn. simpl.

  (* Trivial *)
  reflexivity.
Defined.
\end{lstlisting}

\begin{lstlisting}[language=ML, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
  rewrite (gen n m).
  reflexivity.
Defined.
\end{lstlisting}
